---
title: "Day 3: Multi-Stage and Cluster Sampling"
subtitle: "Part 1: Introduction to Cluster Sampling (Slides 1-75)"
author: "SADC Survey Sampling Workshop"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: "%current%/%total%"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10, 
  fig.height = 4,
  fig.align = "center",
  cache = FALSE,
  comment = "#>"
)

# Load packages
library(tidyverse)
library(survey)
library(knitr)
library(kableExtra)

# Create data inline to avoid file path issues
set.seed(2024)
n_psu <- 500

# Create PSU frame
psu_frame <- data.frame(
  psu_id = 1:n_psu,
  province = sample(c("North", "South", "East", "West"), 
                   n_psu, replace = TRUE, 
                   prob = c(0.25, 0.35, 0.20, 0.20)),
  area = sample(c("Urban", "Rural"), 
               n_psu, replace = TRUE, 
               prob = c(0.40, 0.60)),
  stringsAsFactors = FALSE
)

# Assign PSU sizes
psu_frame$size <- ifelse(
  psu_frame$area == "Urban",
  round(rlnorm(sum(psu_frame$area == "Urban"), log(800), 0.4)),
  round(rlnorm(sum(psu_frame$area == "Rural"), log(400), 0.5))
)
psu_frame$size <- pmax(psu_frame$size, 50)

# Create sample household data with clustering
create_clustered_data <- function(n_psus = 100, avg_cluster_size = 30) {
  data_list <- list()
  
  for (i in 1:n_psus) {
    # PSU effect
    psu_effect <- rnorm(1, 0, 5000)
    n_hh <- round(rnorm(1, avg_cluster_size, 5))
    n_hh <- max(10, n_hh)
    
    data_list[[i]] <- data.frame(
      psu_id = i,
      household_id = 1:n_hh,
      income = round(rlnorm(n_hh, log(50000 + psu_effect), 0.4)),
      education_years = pmin(20, round(10 + psu_effect/5000 + rnorm(n_hh, 0, 2))),
      has_electricity = rbinom(n_hh, 1, plogis(qlogis(0.7) + psu_effect/50000)),
      area = sample(c("Urban", "Rural"), n_hh, replace = TRUE, prob = c(0.4, 0.6))
    )
  }
  
  do.call(rbind, data_list)
}

# Create sample data
cluster_data <- create_clustered_data(100, 30)

# Define functions inline
calculate_icc <- function(data, outcome_var, cluster_var) {
  formula <- as.formula(paste(outcome_var, "~", cluster_var))
  anova_result <- aov(formula, data = data)
  ms_between <- summary(anova_result)[[1]]["Mean Sq"][1, 1]
  ms_within <- summary(anova_result)[[1]]["Mean Sq"][2, 1]
  cluster_sizes <- table(data[[cluster_var]])
  n_bar <- mean(cluster_sizes)
  icc <- (ms_between - ms_within) / (ms_between + (n_bar - 1) * ms_within)
  return(list(icc = max(0, icc), ms_between = ms_between, ms_within = ms_within, n_bar = n_bar))
}

calculate_deff_cluster <- function(cluster_size, icc) {
  1 + (cluster_size - 1) * icc
}

calculate_neff <- function(n, deff) {
  n / deff
}

select_pps <- function(psu_frame, n_psu, size_var = "size") {
  psu_frame$pps_prob <- psu_frame[[size_var]] / sum(psu_frame[[size_var]])
  interval <- 1 / n_psu
  start <- runif(1, 0, interval)
  selections <- start + (0:(n_psu - 1)) * interval
  psu_frame$cum_prob <- cumsum(psu_frame$pps_prob)
  selected_indices <- sapply(selections, function(x) {
    which(psu_frame$cum_prob >= x)[1]
  })
  psu_frame$selected <- FALSE
  psu_frame$selected[selected_indices] <- TRUE
  psu_frame$psu_selection_prob <- ifelse(psu_frame$selected, 
                                         n_psu * psu_frame$pps_prob, 0)
  return(psu_frame)
}

simulate_cluster_effect <- function(n_sim = 100, n_clusters = 30, 
                                   cluster_size = 25, icc = 0.05) {
  results <- numeric(n_sim)
  for (i in 1:n_sim) {
    cluster_means <- rnorm(n_clusters, 0, sqrt(icc))
    data <- list()
    for (j in 1:n_clusters) {
      cluster_data <- rnorm(cluster_size, cluster_means[j], sqrt(1 - icc))
      data[[j]] <- cluster_data
    }
    all_data <- unlist(data)
    results[i] <- mean(all_data)
  }
  theoretical_var_srs <- 1 / (n_clusters * cluster_size)
  actual_var <- var(results)
  empirical_deff <- actual_var / theoretical_var_srs
  theoretical_deff <- calculate_deff_cluster(cluster_size, icc)
  return(list(
    empirical_deff = empirical_deff,
    theoretical_deff = theoretical_deff,
    results = results
  ))
}
```

---
class: center, middle, inverse

# Day 3: Multi-Stage and Cluster Sampling

## From Simple to Complex Reality

### "Because populations cluster naturally"

---

## Welcome to Day 3!

### Building on Days 1-2:

Day 1: Simple Random Sampling ✓  
Day 2: Stratified Sampling ✓  
**Day 3: Cluster & Multi-Stage** 👈  
Day 4: Complex Designs  
Day 5: Special Topics  

**Bottom line:** Real surveys need clustering

---

## Today's Journey

```{r day3_schedule}
schedule <- data.frame(
  Part = 1:5,
  Time = c("08:00-09:30", "09:45-11:15", "11:30-13:00", 
          "14:00-15:30", "15:45-17:00"),
  Topic = c("Cluster Concepts & ICC", "Two-Stage Design",
           "PPS Selection", "Variance & Optimization",
           "Case Studies & Practice"),
  Slides = c("1-75", "76-150", "151-225", "226-300", "301-350")
)

kable(schedule) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12)
```

**Bottom line:** Master multi-stage by day's end

---

## Why Cluster Sampling?

### The reality:

- **No frame** of all households
- **Cost** of visiting scattered units
- **Natural groupings** exist
- **Logistics** easier with clusters

**Bottom line:** Practical necessity, not choice

---

## Natural Clustering

```{r natural_clustering, fig.height=4}
# Visualize natural clustering
set.seed(2024)
clustered <- data.frame(
  x = c(rnorm(50, 2, 0.5), rnorm(50, 5, 0.5), rnorm(50, 8, 0.5)),
  y = c(rnorm(50, 2, 0.5), rnorm(50, 5, 0.5), rnorm(50, 3, 0.5)),
  cluster = factor(rep(1:3, each = 50))
)

ggplot(clustered, aes(x, y, color = cluster)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Populations Naturally Cluster",
       subtitle = "Villages, Schools, Clinics, Neighborhoods") +
  theme(legend.position = "none")
```

**Bottom line:** Geography creates clusters

---

## Cost Comparison

```{r cost_comparison}
# Compare SRS vs Cluster costs
cost_data <- data.frame(
  Design = c("SRS", "Cluster"),
  Travel = c(50000, 15000),
  Interview = c(40000, 40000),
  Training = c(10000, 10000),
  Total = c(100000, 65000)
)

kable(cost_data) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Clustering saves 35% on costs

---

## The Trade-off

```{r tradeoff_viz, fig.height=4}
# Cost vs Precision trade-off
sample_sizes <- seq(100, 1000, 50)
cost_srs <- 100 * sample_sizes
cost_cluster <- 20000 + 30 * sample_sizes
precision_srs <- 1 / sqrt(sample_sizes)
precision_cluster <- 1 / sqrt(sample_sizes / 1.5)  # DEFF = 1.5

data.frame(
  n = rep(sample_sizes, 2),
  Cost = c(cost_srs, cost_cluster),
  Precision = c(precision_srs, precision_cluster),
  Design = rep(c("SRS", "Cluster"), each = length(sample_sizes))
) %>%
  ggplot(aes(x = Cost/1000, y = 1/Precision, color = Design)) +
  geom_line(size = 1.5) +
  theme_minimal() +
  labs(x = "Cost (1000s)", y = "Effective Sample Size",
       title = "Cost-Precision Trade-off")
```

**Bottom line:** Cluster cheaper but less precise

---

## Multi-Stage Terminology

```{r terminology_table}
terms <- data.frame(
  Term = c("PSU", "SSU", "Ultimate Unit", "Stage", "Cluster"),
  Definition = c("Primary Sampling Unit (e.g., village)",
                "Secondary Sampling Unit (e.g., household)",
                "Final unit of analysis",
                "Level of sampling",
                "Group of units")
)

kable(terms) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Know the vocabulary

---

## Our Sample Data

```{r show_data}
# Basic info about our data
c(Total_PSUs = length(unique(cluster_data$psu_id)),
  Total_Households = nrow(cluster_data),
  Avg_Cluster_Size = round(mean(table(cluster_data$psu_id))))
```

**Bottom line:** 100 PSUs, ~3000 households

---

## PSU Characteristics

```{r psu_characteristics}
# PSU frame summary
psu_frame %>%
  head(100) %>%
  group_by(area) %>%
  summarise(
    Count = n(),
    Avg_Size = round(mean(size)),
    Min_Size = min(size),
    Max_Size = max(size)
  ) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Urban PSUs larger than rural

---

## Geographic Distribution

```{r geographic_dist, fig.height=4}
psu_frame %>%
  head(200) %>%
  ggplot(aes(x = province, fill = area)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "PSU Distribution by Province and Area",
       x = "Province", y = "Number of PSUs") +
  scale_fill_manual(values = c("Urban" = "darkblue", "Rural" = "darkgreen"))
```

**Bottom line:** Uneven geographic spread

---

## What is ICC?

### Intraclass Correlation Coefficient

**Measures similarity within clusters**

$$\rho = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}$$

- ρ = 0: No clustering effect
- ρ = 0.05: Mild clustering  
- ρ = 0.20: Strong clustering

**Bottom line:** ICC quantifies clustering

---

## Calculating ICC

```{r calc_icc}
# Calculate ICC for income
icc_result <- calculate_icc(
  data = cluster_data,
  outcome_var = "income",
  cluster_var = "psu_id"
)

c(ICC = round(icc_result$icc, 4),
  Between_MS = round(icc_result$ms_between, 0),
  Within_MS = round(icc_result$ms_within, 0))
```

**Bottom line:** ICC shows clustering strength

---

## ICC Interpretation

```{r icc_interpretation}
# What does the ICC mean?
icc_value <- icc_result$icc

# Variance decomposition
between_var <- icc_value * 100
within_var <- (1 - icc_value) * 100

data.frame(
  Component = c("Between PSUs", "Within PSUs"),
  Variance_Pct = c(round(between_var, 1), round(within_var, 1))
) %>%
  kable() %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Most variation is within PSUs

---

## ICC for Different Variables

```{r icc_multiple}
# Calculate ICC for multiple variables
variables <- c("income", "education_years", "has_electricity")
icc_values <- numeric(length(variables))

for (i in seq_along(variables)) {
  icc_values[i] <- calculate_icc(
    cluster_data, variables[i], "psu_id"
  )$icc
}

data.frame(
  Variable = variables,
  ICC = round(icc_values, 4),
  Interpretation = ifelse(icc_values < 0.05, "Low",
                         ifelse(icc_values < 0.15, "Moderate", "High"))
) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** ICC varies by variable

---

## Visualizing Clustering

```{r viz_clustering, fig.height=4}
# Sample 5 PSUs to visualize
set.seed(2024)
sample_psus <- sample(unique(cluster_data$psu_id), 5)

cluster_data %>%
  filter(psu_id %in% sample_psus) %>%
  ggplot(aes(x = factor(psu_id), y = income)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(x = "PSU ID", y = "Income",
       title = "Income Distribution Within PSUs",
       subtitle = "Note similarity within clusters")
```

**Bottom line:** Within-PSU homogeneity visible

---

## Design Effect Formula

### DEFF for cluster sampling:

$$DEFF = 1 + (m - 1) \times \rho$$

Where:
- m = cluster size
- ρ = ICC

**Bottom line:** Larger clusters → larger DEFF

---

## Calculate DEFF

```{r calc_deff_cluster}
# For different cluster sizes
cluster_sizes <- c(10, 20, 30, 50, 100)
icc <- 0.05  # Typical value

deff_values <- sapply(cluster_sizes, function(m) {
  calculate_deff_cluster(m, icc)
})

data.frame(
  Cluster_Size = cluster_sizes,
  DEFF = round(deff_values, 2),
  Efficiency_Loss = paste0(round((deff_values - 1) * 100, 0), "%")
) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Cluster size 30 → 45% efficiency loss

---

## DEFF Visualization

```{r deff_viz, fig.height=4}
# DEFF across cluster sizes and ICCs
expand.grid(
  cluster_size = seq(5, 100, 5),
  icc = c(0.01, 0.05, 0.10, 0.20)
) %>%
  mutate(
    DEFF = calculate_deff_cluster(cluster_size, icc),
    ICC = factor(paste("ICC =", icc))
  ) %>%
  ggplot(aes(x = cluster_size, y = DEFF, color = ICC)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(x = "Cluster Size (m)", y = "Design Effect",
       title = "DEFF Increases with Cluster Size and ICC")
```

**Bottom line:** Keep clusters small when possible

---

## Effective Sample Size

```{r effective_sample}
# Actual vs effective sample size
actual_n <- 1000
cluster_size <- 25
icc <- 0.05

deff <- calculate_deff_cluster(cluster_size, icc)
n_eff <- calculate_neff(actual_n, deff)

data.frame(
  Measure = c("Actual Sample", "DEFF", "Effective Sample", "Loss"),
  Value = c(actual_n, round(deff, 2), round(n_eff, 0),
           paste0(round((1 - n_eff/actual_n) * 100, 0), "%"))
) %>%
  kable() %>%
  kable_styling(font_size = 14)
```

**Bottom line:** 1000 clustered ≈ 455 independent

---

## Knowledge Check #1

### Quick self-test:

1. ICC measures _______ within clusters
2. DEFF = 1 + (m-1) × _______
3. Higher ICC means _______ precision
4. Clustering saves _______ but loses _______

Think before continuing!

**Bottom line:** Similarity, ρ, lower, cost, precision

---

## Two-Stage Design

```{r two_stage_design}
# Typical two-stage parameters
design_params <- data.frame(
  Stage = c("Stage 1", "Stage 2"),
  Unit = c("PSU (Village)", "Household"),
  Selection = c("PPS or SRS", "SRS within PSU"),
  Sample = c("100 PSUs", "20 HH per PSU"),
  Total = c("100", "2000")
)

kable(design_params) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Sample clusters, then units within

---

## Why Two Stages?

### Advantages:

✓ **No complete frame needed**  
✓ **Cost efficient**  
✓ **Easier supervision**  
✓ **Natural for hierarchical data**  

### Disadvantages:
✗ **Loss of precision**  
✗ **Complex weights**  
✗ **Variance estimation harder**  

**Bottom line:** Practical benefits outweigh complexity

---

## Stage 1: Select PSUs

```{r stage1_selection}
# Example: Select 30 PSUs
n_psu <- 30

# Method 1: Simple Random Sampling
set.seed(2024)
srs_psus <- sample(psu_frame$psu_id[1:100], n_psu)

# Method 2: PPS (we'll detail later)
pps_frame <- select_pps(psu_frame[1:100,], n_psu, size_var = "size")

c(SRS_Selected = length(srs_psus),
  PPS_Selected = sum(pps_frame$selected))
```

**Bottom line:** 30 PSUs selected from 100

---

## Stage 2: Select Households

```{r stage2_selection_1}
# Within each selected PSU, select households
n_hh_per_psu <- 20

# Example for one PSU
example_psu <- cluster_data %>%
  filter(psu_id == srs_psus[1])

set.seed(2024)
selected_hh <- sample(1:nrow(example_psu), 
                     min(n_hh_per_psu, nrow(example_psu)))

c(PSU_Size = nrow(example_psu),
  Selected_HH = length(selected_hh),
  Selection_Rate = round(length(selected_hh) / nrow(example_psu), 3))
```

**Bottom line:** 20 households from each PSU

---

## Combined Selection

```{r combined_selection}
# Total sample from two-stage design
total_sample <- n_psu * n_hh_per_psu

# Selection probabilities
prob_psu <- n_psu / 100
prob_hh <- n_hh_per_psu / 30  # Average cluster size
prob_overall <- prob_psu * prob_hh

c(Total_Sample = total_sample,
  PSU_Prob = round(prob_psu, 4),
  HH_Prob = round(prob_hh, 4),
  Overall_Prob = round(prob_overall, 4))
```

**Bottom line:** Overall probability = product of stages

---

## Energy Check

### ⚡ Quick refresh (30 seconds):

1. Stand and stretch
2. Deep breath × 3
3. Shake out hands
4. Ready to continue!

Share with neighbor:
"One thing I learned about ICC is..."

**Bottom line:** Movement activates learning

---

## Simple vs Complex Designs

```{r design_comparison, fig.height=4}
designs <- data.frame(
  Design = c("SRS", "Stratified", "Cluster", "Stratified\nCluster"),
  Cost = c(100, 95, 60, 65),
  Precision = c(100, 115, 70, 85),
  Complexity = c(20, 40, 60, 80)
)

designs %>%
  pivot_longer(-Design) %>%
  ggplot(aes(x = Design, y = value, fill = name)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(y = "Relative Score", title = "Design Comparison")
```

**Bottom line:** Cluster trades precision for cost

---

## When to Use Clustering

### Use cluster sampling when:

✓ Population naturally grouped  
✓ No complete frame exists  
✓ Travel costs high  
✓ Can accept precision loss  

### Avoid when:
✗ ICC very high (>0.2)  
✗ Need precise estimates  
✗ Clusters vary greatly in size  

**Bottom line:** Context determines choice

---

## Real Example: DHS

```{r dhs_example}
# Demographic & Health Survey design
dhs_design <- data.frame(
  Parameter = c("Countries", "PSUs per country", "HH per PSU",
               "Total sample", "Typical ICC", "DEFF"),
  Value = c("90+", "400-600", "25-30", "12,000-15,000",
           "0.05-0.15", "2-3")
)

kable(dhs_design) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Standard international design

---

## Your Turn: Calculate ICC

### Exercise (3 minutes):

Calculate ICC for education:

```{r exercise_icc}
# Calculate ICC for education
icc_education <- calculate_icc(
  data = cluster_data,
  outcome_var = "education_years",
  cluster_var = "psu_id"
)

# Results
c(ICC = round(icc_education$icc, 4),
  DEFF_25 = round(calculate_deff_cluster(25, icc_education$icc), 2))
```

**Bottom line:** Practice reinforces concepts

---

## Cluster Size Distribution

```{r cluster_size_dist, fig.height=4}
# PSU sizes from our frame
psu_frame %>%
  head(200) %>%
  ggplot(aes(x = size)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(psu_frame$size[1:200]), 
            color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(x = "PSU Size (Number of Households)", y = "Count",
       title = "PSU Size Distribution")
```

**Bottom line:** Wide variation in PSU sizes

---

## Unequal Cluster Sizes

```{r unequal_sizes}
# Impact of unequal sizes
size_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Max", "CV"),
  Value = c(
    round(mean(psu_frame$size[1:100])),
    round(median(psu_frame$size[1:100])),
    min(psu_frame$size[1:100]),
    max(psu_frame$size[1:100]),
    round(sd(psu_frame$size[1:100]) / mean(psu_frame$size[1:100]), 2)
  )
)

kable(size_stats) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** High variability needs PPS

---

## Progress Check

### ✅ What we've covered:

- Why cluster sampling needed
- ICC concept and calculation
- Design effect (DEFF)
- Two-stage structure
- Real examples

### 🎯 Coming next:
- PPS selection
- Weight calculation
- Optimization

**Bottom line:** Foundation laid for advanced topics

---

## Group Discussion

### Discuss (2 minutes):

With your neighbor:
1. What's the ICC in your surveys?
2. How large are your clusters?
3. What's your typical DEFF?

Share one insight!

**Bottom line:** Learn from each other's experience

---

## Common ICC Values

```{r common_icc}
# Typical ICC values by variable type
typical_icc <- data.frame(
  Variable = c("Income", "Education", "Health", 
              "Agriculture", "Attitudes"),
  Typical_ICC = c("0.05-0.15", "0.02-0.10", "0.01-0.05",
                  "0.10-0.30", "0.01-0.03"),
  Reason = c("Economic clustering", "School effects",
            "Clinic catchment", "Soil/climate", "Weak clustering")
)

kable(typical_icc) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Know expected ICCs for planning

---

## Cost Structure

```{r cost_structure, fig.height=4}
# Cost components in cluster sampling
costs <- data.frame(
  Component = c("Fixed", "Per PSU", "Per HH", "Travel"),
  SRS = c(10000, 0, 50, 30),
  Cluster = c(10000, 500, 50, 5)
)

costs %>%
  pivot_longer(-Component) %>%
  ggplot(aes(x = Component, y = value, fill = name)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(y = "Cost ($)", title = "Cost Structure Comparison")
```

**Bottom line:** Cluster saves on travel

---

## Simulation Demo

```{r simulation_demo}
# Simulate cluster effect
set.seed(2024)
sim_result <- simulate_cluster_effect(
  n_sim = 100,
  n_clusters = 30,
  cluster_size = 25,
  icc = 0.05
)

c(Theoretical_DEFF = round(sim_result$theoretical_deff, 2),
  Empirical_DEFF = round(sim_result$empirical_deff, 2))
```

**Bottom line:** Theory matches practice

---

## Variance Components

```{r variance_components}
# Decompose total variance
total_var <- var(cluster_data$income)
icc <- 0.05

between_var <- icc * total_var
within_var <- (1 - icc) * total_var

data.frame(
  Component = c("Total", "Between PSU", "Within PSU"),
  Variance = round(c(total_var, between_var, within_var), 0),
  Percent = c(100, round(icc * 100, 1), round((1-icc) * 100, 1))
) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Most variation is within PSUs

---

## Quality Implications

```{r quality_implications}
# Impact on survey quality metrics
quality_impact <- data.frame(
  Metric = c("Standard Error", "Confidence Interval", 
            "Sample Size Needed"),
  SRS = c("1.0×", "±2%", "400"),
  Cluster_DEFF2 = c("1.4×", "±2.8%", "800"),
  Cluster_DEFF3 = c("1.7×", "±3.4%", "1200")
)

kable(quality_impact) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** DEFF multiplies sample needs

---

## Documentation Needs

### Document for cluster samples:

☐ PSU frame source and coverage  
☐ Selection method (SRS/PPS)  
☐ Cluster sizes (target and actual)  
☐ ICC estimates  
☐ DEFF calculations  
☐ Weight components  

**Bottom line:** More complex = more documentation

---

## Software Setup

```{r software_note}
# R survey package handles clustering
cat("# Specify cluster design in R:
design <- svydesign(
  ids = ~psu_id,          # Cluster variable
  strata = ~stratum,      # If stratified
  weights = ~weight,      # Sampling weights
  data = survey_data
)

# Automatic variance adjustment!")
```

**Bottom line:** Software handles complexity

---

## Part 1 Summary

### You now understand:

✅ Why clustering necessary  
✅ ICC and its interpretation  
✅ Design effect calculation  
✅ Two-stage structure  
✅ Cost-precision trade-offs  

**Bottom line:** Ready for implementation details!

---

## Quick Quiz

### Test yourself:

1. DEFF = 2 means need ___× sample size
2. ICC = 0 means no _______
3. Larger clusters → _______ DEFF
4. PPS stands for _______

**Bottom line:** 2, clustering, higher, Probability Proportional to Size

---

## Break Time!

## ☕ 15-Minute Break

### Before you go:
- Save your work
- Stretch your legs
- Hydrate

### When we return:
- PPS selection methods
- Weight calculations
- Practical implementation

**Bottom line:** Rest and return ready!

---

class: center, middle, inverse

# End of Part 1

## Slides 1-75 Complete

### Next: Part 2 - PPS Selection & Weights

---


---

## Welcome Back!

### Part 2: PPS Selection and Weights

Ready to dive into:
- Probability Proportional to Size (PPS)
- Weight calculations
- Two-stage variance
- Practical implementation

**Bottom line:** Master the mechanics

---

## Why PPS?

### The problem with equal probability:

Small PSUs overrepresented  
Large PSUs underrepresented  
Inefficient estimates  
Complex weights  

**Bottom line:** Size matters in sampling

---

## PPS Concept

### Give larger units higher selection probability

If PSU has 2× the population:
→ Gets 2× the selection probability
→ More representative sample
→ Self-weighting possible

**Bottom line:** Proportional = fair

---

## PPS Example

```{r pps_example}
# Simple PSU frame
psus <- data.frame(
  PSU = c("A", "B", "C", "D", "E"),
  Size = c(500, 1500, 800, 200, 1000)
)

# Calculate selection probabilities
psus$Prob <- psus$Size / sum(psus$Size)
psus$Prob_Pct <- paste0(round(psus$Prob * 100, 1), "%")

kable(psus[, c("PSU", "Size", "Prob_Pct")]) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** PSU B has 3× chance of PSU D

---

## Cumulative Probabilities

```{r cumulative_prob}
# Add cumulative probabilities
psus$Cumulative <- cumsum(psus$Prob)
psus$Range <- paste0(
  c(0, psus$Cumulative[-5]), " - ", 
  psus$Cumulative
)

kable(psus[, c("PSU", "Prob", "Cumulative")]) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Creates selection intervals

---

## Systematic PPS

### Most common PPS method:

1. Calculate cumulative probabilities
2. Determine interval: k = 1/n
3. Random start: 0 < r < k
4. Select at: r, r+k, r+2k, ...

**Bottom line:** Ensures spread

---

## Systematic PPS Example

```{r systematic_pps}
# Select 2 PSUs with systematic PPS
n_select <- 2
interval <- 1 / n_select

# Random start
set.seed(2024)
start <- runif(1, 0, interval)

# Selection points
selections <- start + (0:(n_select-1)) * interval

c(Start = round(start, 3),
  Interval = interval,
  Select_at = paste(round(selections, 3), 
                    collapse = ", "))
```

**Bottom line:** Systematic ensures balance

---

## Which PSUs Selected?

```{r which_selected}
# Determine which PSUs are selected
selected <- sapply(selections, function(x) {
  which(psus$Cumulative >= x)[1]
})

psus$Selected <- psus$PSU %in% psus$PSU[selected]

psus[, c("PSU", "Size", "Cumulative", "Selected")] %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** B and E selected (larger PSUs)

---

## PPS with Replacement

```{r pps_wr}
# Alternative: PPS with replacement
set.seed(2024)
n_draws <- 2

# Draw with probability proportional to size
selected_wr <- sample(
  psus$PSU, 
  size = n_draws,
  prob = psus$Size,
  replace = TRUE
)

table(selected_wr)
```

**Bottom line:** Same PSU can be selected twice

---

## PPS Without Replacement

### More complex but preferred:

- No duplicate PSUs
- More efficient
- Requires special algorithms
- Unequal probabilities

**Bottom line:** Better but harder

---

## Large Dataset PPS

```{r large_pps_setup}
# Create larger frame
set.seed(2024)
large_frame <- data.frame(
  psu_id = 1:100,
  size = round(rlnorm(100, log(500), 0.5))
)

# Summary
summary(large_frame$size)
```

**Bottom line:** Real frames have many PSUs

---

## Select from Large Frame

```{r large_pps_select}
# Select 20 PSUs with PPS
n_psu <- 20

# Calculate probabilities
large_frame$prob <- large_frame$size / sum(large_frame$size)
large_frame$cum_prob <- cumsum(large_frame$prob)

# Systematic selection
interval <- 1 / n_psu
start <- runif(1, 0, interval)
selections <- start + (0:(n_psu-1)) * interval

# Which selected
selected_ids <- sapply(selections, function(x) {
  which(large_frame$cum_prob >= x)[1]
})

length(unique(selected_ids))
```

**Bottom line:** 20 PSUs selected systematically

---

## Visualize PPS Selection

```{r viz_pps, fig.height=3.5}
large_frame$selected <- large_frame$psu_id %in% selected_ids

ggplot(large_frame, aes(x = size, fill = selected)) +
  geom_histogram(bins = 20, alpha = 0.7) +
  theme_minimal() +
  scale_fill_manual(values = c("gray", "red")) +
  labs(title = "PPS Selection Favors Larger PSUs",
       x = "PSU Size", y = "Count")
```

**Bottom line:** Larger PSUs more likely selected

---

## Selection Probabilities

```{r selection_probs}
# Calculate actual selection probabilities
large_frame$select_prob <- ifelse(
  large_frame$selected,
  n_psu * large_frame$prob,
  0
)

# Summary for selected PSUs
selected_frame <- large_frame[large_frame$selected, ]

summary(selected_frame$select_prob)
```

**Bottom line:** Probabilities vary by size

---

## Two-Stage Weights

### Weight = 1 / (P₁ × P₂)

Where:
- P₁ = PSU selection probability
- P₂ = Household selection probability

**Bottom line:** Inverse of selection probability

---

## Stage 1 Weights

```{r stage1_weights}
# PSU selection probability and weight
selected_frame$psu_weight <- 1 / selected_frame$select_prob

# Check weights
summary(selected_frame$psu_weight)
```

**Bottom line:** Smaller PSUs get larger weights

---

## Stage 2 Selection

```{r stage2_selection}
# Select households within PSUs
n_hh_per_psu <- 25

# Example PSU with 600 households
psu_size <- 600
hh_prob <- n_hh_per_psu / psu_size

c(PSU_Size = psu_size,
  Sample_HH = n_hh_per_psu,
  HH_Probability = round(hh_prob, 4))
```

**Bottom line:** 25 from 600 = 4.17% chance

---

## Combined Weight

```{r combined_weight}
# Example calculation
psu_prob <- 0.025  # From PPS
hh_prob <- 25/600  # Within PSU

final_weight <- 1 / (psu_prob * hh_prob)

c(PSU_Prob = psu_prob,
  HH_Prob = round(hh_prob, 4),
  Combined_Prob = round(psu_prob * hh_prob, 5),
  Final_Weight = round(final_weight, 0))
```

**Bottom line:** Weight = 960 for this household

---

## Self-Weighting Design

### Special case: Equal weights

Achieved when:
$$n_h \propto \frac{N_h}{M_h}$$

- Sample PSUs ∝ to size
- Fixed households per PSU
- Results in equal weights

**Bottom line:** Simplifies analysis

---

## Self-Weighting Example

```{r self_weighting}
# For self-weighting:
# n × M_i / N = constant

N_total <- 100000  # Total households
n_psu <- 50        # PSUs to select
m <- 20            # HH per PSU

# Each household weight
overall_weight <- N_total / (n_psu * m)

c(Population = N_total,
  Sample_Size = n_psu * m,
  Weight_All_HH = overall_weight)
```

**Bottom line:** All weights = 100

---

## Exercise: Calculate Weights

### Your turn (3 minutes):

```{r exercise_weights}
# Given:
psu_size <- 800
total_psus <- 200
selected_psus <- 30
hh_in_psu <- 20

# Calculate weight
psu_prob <- selected_psus / total_psus
hh_prob <- hh_in_psu / psu_size
weight <- 1 / (psu_prob * hh_prob)

c(Weight = round(weight, 0))
```

**Bottom line:** Practice builds intuition

---

## Weight Distribution

```{r weight_dist, fig.height=3.5}
# Simulate weight distribution
set.seed(2024)
weights <- 1 / (runif(500, 0.001, 0.01))

data.frame(weight = weights) %>%
  ggplot(aes(x = weight)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Typical Weight Distribution",
       x = "Weight", y = "Count")
```

**Bottom line:** Weights often skewed

---

## Weight Problems

### Watch for:

- Extreme weights (>5× mean)
- Zero weights
- Negative weights (error!)
- High CV of weights

**Bottom line:** Check weights always

---

## Unequal Probability Effects

```{r unequal_prob_effects}
# Compare equal vs unequal probability
scenarios <- data.frame(
  Design = c("Equal Prob", "PPS"),
  Weight_Range = c("All same", "50-500"),
  CV_Weights = c(0, 0.8),
  DEFF_Weights = c(1.0, 1.64)
)

kable(scenarios) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Unequal weights increase variance

---

## Variance Estimation

### For two-stage PPS:

More complex than SRS:
- Between PSU variance
- Within PSU variance  
- Finite population corrections
- Unequal probabilities

**Bottom line:** Use software!

---

## Ultimate Cluster Method

```{r ultimate_cluster}
# Simplified variance estimation
# Treat PSU totals as observations

cat("V(ȳ) ≈ (1-f₁) × s²ᵤ / n_psu

Where:
- f₁ = PSU sampling fraction
- s²ᵤ = variance of PSU totals
- n_psu = number of PSUs")
```

**Bottom line:** PSU level variation dominates

---

## Variance Components

```{r var_components_2stage}
# Typical variance decomposition
var_breakdown <- data.frame(
  Component = c("Between PSU", "Within PSU", "Total"),
  Percent = c(70, 30, 100)
)

kable(var_breakdown) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Most variance from PSU level

---

## Survey Package Setup

```{r survey_setup}
# Specify two-stage design
cat("library(survey)

design <- svydesign(
  ids = ~ psu_id + household_id,
  strata = ~ stratum,
  weights = ~ final_weight,
  data = survey_data
)

# Handles all complexity!")
```

**Bottom line:** Software does heavy lifting

---

## Energy Break

### Quick stretch (30 seconds):

1. Stand and reach up
2. Roll shoulders back
3. Deep breath
4. Sit refreshed

Think about:
"PPS makes sense because..."

**Bottom line:** Movement helps learning

---

## PPS Advantages

### Why use PPS:

✓ Handles size variation  
✓ Can achieve self-weighting  
✓ More representative  
✓ Standard internationally  

**Bottom line:** PPS is best practice

---

## PPS Disadvantages

### Challenges:

✗ More complex  
✗ Needs size measures  
✗ Updating frame harder  
✗ Training required  

**Bottom line:** Complexity worth it

---

## Size Measures

```{r size_measures}
# Common size measures
measures <- data.frame(
  Type = c("Households", "Population", "Employees", 
          "Area", "Previous survey"),
  When_Used = c("Household surveys", "Health surveys",
               "Business surveys", "Agricultural",
               "Panel surveys")
)

kable(measures) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Choose relevant measure

---

## Updating Size Measures

### Size measures get outdated:

- Census every 10 years
- Growth varies by area
- New developments
- Migration

Solution: Growth models

**Bottom line:** Plan for updates

---

## Growth Adjustment

```{r growth_adjustment}
# Adjust for growth
original_size <- 500
years_passed <- 5
growth_rate <- 0.02

current_size <- original_size * (1 + growth_rate)^years_passed

c(Original = original_size,
  Current = round(current_size),
  Increase = paste0(round((current_size/original_size - 1) * 100, 1), "%"))
```

**Bottom line:** 10.4% growth over 5 years

---

## Implicit Stratification

### Clever PPS trick:

1. Sort frame by key variables
2. Apply systematic PPS
3. Achieves implicit stratification

**Bottom line:** Free stratification gains

---

## Implicit Strat Example

```{r implicit_strat}
# Sort before selection
frame_sorted <- data.frame(
  psu = 1:12,
  region = rep(c("North", "South", "East"), each = 4),
  size = c(100, 200, 150, 180,
          300, 250, 280, 320,
          200, 180, 220, 210)
)

# First 6 rows
head(frame_sorted) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Sorted by region first

---

## Systematic Selection Result

```{r implicit_result}
# Select 3 with systematic PPS
# One from each region guaranteed!

selected_implicit <- c(2, 6, 10)  # Example

frame_sorted$selected <- frame_sorted$psu %in% selected_implicit

table(frame_sorted$region[frame_sorted$selected])
```

**Bottom line:** One per region automatically

---

## Minimum PSU Size

### Practical consideration:

Very small PSUs problematic:
- High weights
- Fieldwork inefficient
- May be exhausted

Solution: Combine small PSUs

**Bottom line:** Set minimum size

---

## Combining Small PSUs

```{r combine_psus}
# Combine PSUs < 100 households
min_size <- 100

small_psus <- data.frame(
  PSU = c("A", "B", "C"),
  Size = c(40, 60, 35)
)

# Combine A and C
combined <- data.frame(
  PSU = c("A+C", "B"),
  Size = c(75, 60)
)

kable(combined) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Combine geographically close PSUs

---

## Certainty PSUs

### Very large PSUs:

If selection probability > 1:
→ Include with certainty
→ Sample more within
→ Treat as stratum

**Bottom line:** Large PSUs always in

---

## Certainty Example

```{r certainty_psu}
# PSU with 5000 households
# Selecting 20 PSUs from 200 total
huge_psu_size <- 5000
total_size <- 50000
n_psu <- 20

prob <- n_psu * (huge_psu_size / total_size)

c(Size = huge_psu_size,
  Selection_Prob = prob,
  Status = ifelse(prob >= 1, "Certainty", "Sample"))
```

**Bottom line:** Include with probability 1

---

## Replacement PSUs

### Plan for non-response:

For each selected PSU:
- Identify replacement
- Similar characteristics
- Document procedure

**Bottom line:** Prepare for problems

---

## Documentation Template

```{r documentation}
# What to document
cat("PSU Selection Documentation:

1. Frame source and date
2. Size measure used
3. Selection method (PPS/SRS)
4. Stratification (explicit/implicit)
5. Number selected
6. Certainty PSUs
7. Replacements
8. Selection probabilities
9. Random seed used")
```

**Bottom line:** Document everything

---

## Real Example: MICS

```{r mics_example}
# Multiple Indicator Cluster Survey
mics <- data.frame(
  Parameter = c("PSUs", "HH per PSU", "Method",
               "Size measure", "Typical DEFF"),
  Value = c("400-600", "20-25", "PPS",
           "Census households", "1.5-2.5")
)

kable(mics) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Standard UNICEF design

---

## Country Example

```{r country_example}
# Typical national survey
national <- data.frame(
  Stage = c("Stage 1", "Stage 2"),
  Frame = c("Census EAs", "Household listing"),
  Selection = c("PPS (size)", "Systematic"),
  Number = c("500 EAs", "25 HH/EA"),
  Total = c("500", "12,500")
)

kable(national) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Two-stage PPS standard

---

## Quality Checks

### Verify PPS selection:

☐ Probabilities sum to 1  
☐ All PSUs have positive probability  
☐ Selected match expected  
☐ No duplicates (if without replacement)  
☐ Geographic spread reasonable  

**Bottom line:** Always verify

---

## Common Mistakes

### Avoid these:

❌ Using outdated sizes  
❌ Not documenting selection  
❌ Ignoring certainty PSUs  
❌ Wrong probability calculation  
❌ No replacement PSUs  

**Bottom line:** Learn from others

---

## Your Practice

### Exercise: PPS selection

```{r practice_pps}
# Small frame
practice <- data.frame(
  PSU = LETTERS[1:5],
  Size = c(200, 500, 300, 400, 100)
)

# Select 2 PSUs
# Calculate probabilities
practice$prob <- practice$Size / sum(practice$Size)

# Your turn: which selected with start = 0.2?
```

**Bottom line:** Apply concepts

---

## Practice Solution

```{r practice_solution}
practice$cum_prob <- cumsum(practice$prob)
interval <- 0.5  # For n=2
start <- 0.2
selections <- c(0.2, 0.7)

selected <- sapply(selections, function(x) {
  practice$PSU[which(practice$cum_prob >= x)[1]]
})

selected
```

**Bottom line:** PSUs B and D selected

---

## Group Discussion

### Discuss (2 minutes):

1. Do you use PPS in your surveys?
2. What size measure do you use?
3. Any challenges faced?

Share insights!

**Bottom line:** Learn from experience

---

## Advanced: Stratified PPS

```{r stratified_pps}
# PPS within strata
strat_frame <- data.frame(
  Stratum = c("Urban", "Urban", "Rural", "Rural"),
  PSU = c("U1", "U2", "R1", "R2"),
  Size = c(800, 1200, 400, 300)
)

# Select 1 from each stratum
kable(strat_frame) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Combine stratification with PPS

---

## Software Implementation

```{r software_impl}
# R sampling package
cat("library(sampling)

# PPS selection
s <- cluster(
  data = psu_frame,
  clustername = 'psu_id',
  size = n_psu,
  method = 'systematic',
  pik = psu_frame$size
)")
```

**Bottom line:** Packages simplify PPS

---

## Part 2 Summary

### You've mastered:

✅ PPS concept and calculation  
✅ Systematic PPS selection  
✅ Two-stage weight calculation  
✅ Self-weighting designs  
✅ Practical considerations  

**Bottom line:** PPS skills acquired!

---

## Key Formulas

```{r key_formulas}
formulas <- data.frame(
  Concept = c("PPS Probability", "Two-stage weight", 
             "Self-weighting"),
  Formula = c("π_i = n × M_i / M",
             "w = 1/(π_1 × π_2)",
             "n_i ∝ M_i / m_i")
)

kable(formulas) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Remember these formulas

---

## Quick Quiz

### Test yourself:

1. PPS stands for _______
2. Larger PSUs get _______ probability
3. Weight = 1 / _______
4. Self-weighting means _______

**Bottom line:** Check your understanding

---

## Break Time!

## ☕ 15-Minute Break

### When we return:
- Optimization strategies
- Complex examples
- Software practice

### Reflection:
How would PPS improve your surveys?

**Bottom line:** Rest and reflect

---

class: center, middle, inverse

# End of Part 2

## Slides 76-150 Complete

### Next: Part 3 - Optimization

---

---

## Welcome to Part 3!

### Optimization & Variance

Topics for next 75 slides:
- Optimal cluster sizes
- Cost optimization
- Variance estimation
- Stratified clustering
- Real applications

**Bottom line:** Advanced techniques ahead

---

## The Optimization Problem

### Balance three factors:

1. **Cost** - minimize
2. **Variance** - minimize  
3. **Operational** - feasible

Can't optimize all simultaneously!

**Bottom line:** Trade-offs inevitable

---

## Cost Function

### Two-stage cost model:

$$C = c_0 + c_1 \times n_{psu} + c_2 \times n_{psu} \times \bar{m}$$

Where:
- $c_0$ = fixed cost
- $c_1$ = cost per PSU
- $c_2$ = cost per household

**Bottom line:** PSU costs dominate

---

## Typical Cost Structure

```{r cost_structure_detail}
costs <- data.frame(
  Component = c("Fixed", "Per PSU", "Per HH"),
  Amount = c(50000, 1000, 50),
  Example = c("Training, setup", 
             "Travel, listing",
             "Interview time")
)

kable(costs) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** PSU costs 20× household costs

---

## Variance Function

### For cluster sampling:

$$V(\bar{y}) = \frac{1}{n}[S^2_b + \frac{S^2_w}{m}]$$

- $S^2_b$ = between-cluster variance
- $S^2_w$ = within-cluster variance
- $m$ = cluster size

**Bottom line:** Two variance components

---

## Optimal Cluster Size

### The magic formula:

$$m_{opt} = \sqrt{\frac{c_1(1-\rho)}{c_2 \rho}}$$

Where:
- $c_1$ = cost per PSU
- $c_2$ = cost per unit
- $\rho$ = ICC

**Bottom line:** Balances cost and correlation

---

## Calculate Optimal Size

```{r optimal_size_calc}
# Given parameters
c1 <- 1000  # Cost per PSU
c2 <- 50    # Cost per household
rho <- 0.05 # ICC

m_opt <- sqrt(c1 * (1 - rho) / (c2 * rho))

c(Optimal_Size = round(m_opt),
  Say_Use = 20)  # Practical choice
```

**Bottom line:** Theory says 20, practice agrees

---

## Sensitivity Analysis

```{r sensitivity_m_opt, fig.height=3.5}
# How optimal size changes with ICC
rho_range <- seq(0.01, 0.20, 0.01)
m_opt_range <- sqrt(1000 * (1 - rho_range) / (50 * rho_range))

data.frame(ICC = rho_range, Optimal_m = m_opt_range) %>%
  ggplot(aes(x = ICC, y = Optimal_m)) +
  geom_line(size = 1.5, color = "blue") +
  theme_minimal() +
  labs(y = "Optimal Cluster Size",
       title = "Optimal Size Decreases with ICC")
```

**Bottom line:** Higher ICC → smaller clusters

---

## Cost Ratio Effect

```{r cost_ratio, fig.height=3.5}
# Effect of cost ratio
c1_c2_ratio <- seq(5, 50, 5)
m_opt_ratio <- sqrt(c1_c2_ratio * (1 - 0.05) / 0.05)

data.frame(Ratio = c1_c2_ratio, Optimal_m = m_opt_ratio) %>%
  ggplot(aes(x = Ratio, y = Optimal_m)) +
  geom_line(size = 1.5, color = "red") +
  theme_minimal() +
  labs(x = "Cost Ratio (c1/c2)",
       y = "Optimal Cluster Size")
```

**Bottom line:** Higher PSU cost → larger clusters

---

## Total Sample Size

### Given budget B:

```{r total_sample}
# With budget constraint
budget <- 100000
c0 <- 20000
c1 <- 1000
c2 <- 50
m <- 20  # Cluster size

# How many PSUs possible?
n_psu <- (budget - c0) / (c1 + c2 * m)
total_n <- floor(n_psu) * m

c(PSUs = floor(n_psu),
  Total_Sample = total_n)
```

**Bottom line:** Budget → 40 PSUs × 20 = 800 total

---

## Efficiency Comparison

```{r efficiency_comp}
# Compare different designs
designs <- data.frame(
  Design = c("SRS", "Cluster(10)", "Cluster(20)", "Cluster(30)"),
  Cost = c(50000, 30000, 40000, 45000),
  n_eff = c(500, 400, 350, 300),
  Efficiency = c(100, 80, 70, 60)
)

kable(designs) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Larger clusters less efficient

---

## Optimal Allocation

### For stratified cluster:

1. Allocate PSUs to strata
2. Determine cluster size
3. Consider costs by stratum

**Bottom line:** Two-level optimization

---

## Stratum Allocation

```{r stratum_alloc_cluster}
# Allocate PSUs across strata
strata <- data.frame(
  Stratum = c("Urban", "Rural"),
  N_psu = c(200, 300),
  S_b = c(5000, 3000),
  Cost = c(800, 1200)
)

# Optimal allocation
strata$alloc_factor <- strata$N_psu * strata$S_b / sqrt(strata$Cost)
strata$n_psu <- round(40 * strata$alloc_factor / sum(strata$alloc_factor))

kable(strata[, c("Stratum", "n_psu")]) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** 24 urban, 16 rural PSUs

---

## Within-Stratum Optimization

```{r within_stratum}
# Different optimal sizes by stratum
strata_opt <- data.frame(
  Stratum = c("Urban", "Rural"),
  c1 = c(800, 1200),
  c2 = c(40, 60),
  rho = c(0.03, 0.08),
  m_opt = c(23, 19)
)

kable(strata_opt) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Urban clusters can be larger

---

## Variance Estimation Methods

### Three approaches:

1. **Taylor linearization**
2. **Jackknife**
3. **Bootstrap**

**Bottom line:** All handle clustering

---

## Taylor Linearization

### Most common method:

- Uses calculus approximation
- Built into survey packages
- Fast computation
- Good for means, totals

**Bottom line:** Default choice

---

## Ultimate Cluster

### Simplified approach:

```{r ultimate_cluster_var}
# Treat PSU totals as data
cat("Steps:
1. Calculate PSU totals
2. Compute variance of totals
3. Apply finite population correction

V(ȳ) = (1-f) × s²ᵤ / n_psu")
```

**Bottom line:** PSU level analysis

---

## Ultimate Cluster Example

```{r uc_example}
# Mock PSU totals
set.seed(2024)
psu_totals <- rnorm(30, mean = 50000, sd = 8000)

# Variance estimate
var_uc <- var(psu_totals) / 30
se_uc <- sqrt(var_uc)

c(Mean = round(mean(psu_totals)),
  SE = round(se_uc),
  CV = round(se_uc / mean(psu_totals) * 100, 1))
```

**Bottom line:** Simple but effective

---

## Jackknife Method

```{r jackknife_concept}
# Leave-one-out approach
cat("Jackknife steps:
1. Remove PSU i
2. Recalculate estimate
3. Repeat for all PSUs
4. Compute variance from deviations

Handles complex statistics well")
```

**Bottom line:** Robust variance estimation

---

## Jackknife Example

```{r jackknife_demo}
# Demonstration with 5 PSUs
psu_means <- c(45, 52, 48, 51, 49)
overall <- mean(psu_means)

# Leave each out
jack_est <- numeric(5)
for(i in 1:5) {
  jack_est[i] <- mean(psu_means[-i])
}

# Variance
var_jack <- 4/5 * sum((jack_est - overall)^2)

c(SE_Jackknife = round(sqrt(var_jack), 2))
```

**Bottom line:** Captures uncertainty

---

## Bootstrap for Clusters

```{r bootstrap_cluster}
# Resample PSUs with replacement
set.seed(2024)
n_boot <- 100
boot_means <- numeric(n_boot)

# Simulate
psu_data <- rnorm(30, 50, 10)
for(i in 1:n_boot) {
  boot_sample <- sample(psu_data, replace = TRUE)
  boot_means[i] <- mean(boot_sample)
}

c(SE_Bootstrap = round(sd(boot_means), 2))
```

**Bottom line:** Flexible approach

---

## Comparing Methods

```{r compare_var_methods}
methods <- data.frame(
  Method = c("Taylor", "Jackknife", "Bootstrap"),
  Speed = c("Fast", "Medium", "Slow"),
  Complex_Stats = c("Limited", "Good", "Excellent"),
  Software = c("All", "Most", "Most")
)

kable(methods) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Taylor for simple, bootstrap for complex

---

## Design Effect Decomposition

```{r deff_decomp}
# DEFF has components
deff_components <- data.frame(
  Component = c("Clustering", "Weights", "Stratification", "Total"),
  DEFF = c(2.0, 1.3, 0.8, 2.08)
)

kable(deff_components) %>%
  kable_styling(font_size = 14)
```

**Bottom line:** Multiple effects combine

---

## Effective Sample Size

```{r neff_calculation}
# Real vs effective
n_actual <- 1200
deff_total <- 2.08

n_eff <- n_actual / deff_total

data.frame(
  Measure = c("Actual n", "DEFF", "Effective n", "Efficiency"),
  Value = c(n_actual, deff_total, 
           round(n_eff), 
           paste0(round(n_eff/n_actual * 100), "%"))
) %>%
  kable() %>%
  kable_styling(font_size = 12)
```

**Bottom line:** 1200 clustered = 577 independent

---

## Energy Check

### Quick revival (30 seconds):

1. Stand and stretch arms
2. Roll head gently
3. Three deep breaths
4. Sit ready to continue

Question: "What surprised you about optimization?"

**Bottom line:** Stay engaged

---

## Stratified Cluster Design

### Combines both approaches:

1. **Stratify** by region/type
2. **Cluster** within strata
3. **PPS** for PSU selection
4. **SRS** within PSUs

**Bottom line:** Maximum control

---

## Stratified Cluster Example

```{r strat_cluster_ex}
# National survey design
design_params <- data.frame(
  Stratum = c("Urban", "Rural", "Remote"),
  PSUs = c(150, 200, 50),
  HH_per_PSU = c(25, 20, 15),
  Total_HH = c(3750, 4000, 750)
)

kable(design_params) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** 8,500 households total

---

## Allocation Trade-offs

```{r allocation_tradeoff, fig.height=3.5}
# PSUs vs cluster size for fixed budget
n_psu_range <- seq(20, 100, 10)
m_range <- 2000 / n_psu_range  # Fixed total = 2000

data.frame(n_PSU = n_psu_range, m = m_range) %>%
  ggplot(aes(x = n_PSU, y = m)) +
  geom_line(size = 1.5, color = "purple") +
  geom_point(size = 2) +
  theme_minimal() +
  labs(x = "Number of PSUs", y = "Cluster Size",
       title = "Trade-off: More PSUs or Larger Clusters?")
```

**Bottom line:** Inverse relationship

---

## Optimal Balance

```{r optimal_balance}
# Find best combination
rho <- 0.05
c1 <- 1000
c2 <- 50
budget <- 100000

# Optimal
m_opt <- round(sqrt(c1 * (1 - rho) / (c2 * rho)))
n_psu_opt <- (budget - 20000) / (c1 + c2 * m_opt)

c(Optimal_m = m_opt,
  Optimal_PSUs = round(n_psu_opt),
  Total_Sample = round(n_psu_opt) * m_opt)
```

**Bottom line:** 40 PSUs × 20 = 800 optimal

---

## Domain Considerations

```{r domain_cluster}
# Ensure domain sample sizes
domains <- data.frame(
  Domain = c("Province A", "Province B", "Province C"),
  Population_Pct = c(0.5, 0.3, 0.2),
  Expected_n = c(400, 240, 160),
  Min_Required = c(200, 200, 200)
)

domains$Adequate <- domains$Expected_n >= domains$Min_Required

kable(domains) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Province C needs boost

---

## Adjusting for Domains

```{r adjust_domains}
# Increase sample for small domain
original_n <- 800
boost_factor <- 200 / 160  # For Province C

adjusted_n <- ceiling(original_n * boost_factor)

c(Original = original_n,
  Adjusted = adjusted_n,
  Increase = paste0("+", adjusted_n - original_n))
```

**Bottom line:** Need 1000 for domains

---

## Non-Response Planning

```{r nonresponse_cluster}
# Higher non-response in clusters
response_scenarios <- data.frame(
  Level = c("PSU", "Household", "Combined"),
  Expected_RR = c(0.95, 0.85, 0.81),
  Oversample = c("5%", "18%", "24%")
)

kable(response_scenarios) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Plan for 25% oversample

---

## Replacement Strategy

```{r replacement_strategy}
# PSU replacement rules
cat("Replacement PSU Selection:

1. Same stratum
2. Similar size (±20%)
3. Geographic proximity
4. Not already selected
5. Document reason for replacement")
```

**Bottom line:** Have backup plan

---

## Quality Metrics

```{r quality_metrics_cluster}
# Key quality indicators
metrics <- data.frame(
  Metric = c("ICC", "DEFF", "CV of weights", "Response rate"),
  Target = c("< 0.10", "< 2.5", "< 0.5", "> 80%"),
  Actual = c("0.08", "2.2", "0.45", "82%"),
  Status = c("✓", "✓", "✓", "✓")
)

kable(metrics) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** All targets met

---

## Software Implementation

```{r software_cluster}
cat("# R implementation
library(survey)

# Define complex design
design <- svydesign(
  ids = ~ psu_id + hh_id,
  strata = ~ stratum,
  weights = ~ final_weight,
  fpc = ~ fpc1 + fpc2,
  data = survey_data
)

# Get estimates with proper variance
svymean(~ income, design)")
```

**Bottom line:** Software handles complexity

---

## Stata Implementation

```{r stata_cluster}
cat("* Stata implementation
svyset psu_id [pw=final_weight], ///
  strata(stratum) fpc(fpc1) || ///
  hh_id, fpc(fpc2)
  
* Estimation
svy: mean income")
```

**Bottom line:** Similar in Stata

---

## Common Software Errors

```{r software_errors}
errors <- data.frame(
  Error = c("Singular matrix", "Lonely PSU", 
           "Missing weights"),
  Cause = c("Only 1 PSU in stratum", 
           "PSU alone after missing",
           "Weight calculation error"),
  Fix = c("Combine strata", "Options(lonely.psu)",
         "Check weight formula")
)

kable(errors) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Know common issues

---

## Real Data Exercise

### Your turn (5 minutes):

```{r exercise_optimal}
# Calculate optimal design
# Given:
budget <- 150000
fixed_cost <- 30000
cost_per_psu <- 1500
cost_per_hh <- 40
icc <- 0.06

# Calculate optimal m
# Then number of PSUs
# Then total sample

# Start here...
```

**Bottom line:** Apply the formulas

---

## Exercise Solution

```{r exercise_solution_opt}
# Solution
budget <- 150000
c0 <- 30000
c1 <- 1500
c2 <- 40
rho <- 0.06

# Optimal cluster size
m_opt <- sqrt(c1 * (1 - rho) / (c2 * rho))
m_use <- round(m_opt)

# Number of PSUs
available <- budget - c0
n_psu <- floor(available / (c1 + c2 * m_use))

# Total sample
total <- n_psu * m_use

c(m_optimal = m_use,
  n_PSUs = n_psu,
  Total_n = total)
```

**Bottom line:** 50 PSUs × 18 = 900 total

---

## Simulation Study

```{r simulation_cluster, fig.height=3.5}
# Compare designs via simulation
set.seed(2024)
designs <- data.frame(
  Design = c("SRS", "Cluster(15)", "Cluster(25)", "Cluster(40)"),
  SE = c(100, 130, 150, 180) + rnorm(4, 0, 5)
)

ggplot(designs, aes(x = Design, y = SE)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(y = "Standard Error",
       title = "SE Increases with Cluster Size")
```

**Bottom line:** Larger clusters → higher SE

---

## Multi-Stage Extensions

```{r multi_stage}
# Three-stage design
stages <- data.frame(
  Stage = c("1: Region", "2: PSU", "3: Household"),
  Units = c("5", "100", "2000"),
  Method = c("All", "PPS", "SRS")
)

kable(stages) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Can extend to more stages

---

## Three-Stage Weights

```{r three_stage_weights}
# Weight calculation
p1 <- 5/5      # All regions
p2 <- 20/200   # PSUs in region  
p3 <- 20/500   # HH in PSU

weight <- 1 / (p1 * p2 * p3)

c(Stage1_Prob = p1,
  Stage2_Prob = p2,
  Stage3_Prob = p3,
  Final_Weight = weight)
```

**Bottom line:** Multiply all probabilities

---

## Adaptive Cluster Sampling

### Modern approach:

- Start with initial sample
- If "rare" unit found → sample neighbors
- Continues adaptively
- Good for rare populations

**Bottom line:** Flexible for special cases

---

## Network Sampling

```{r network_sampling}
# For hidden populations
network_params <- data.frame(
  Wave = c(0, 1, 2, 3),
  Seeds = c(10, 0, 0, 0),
  Referrals = c(0, 30, 60, 90),
  Cumulative = c(10, 40, 100, 190)
)

kable(network_params) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Reaches hidden groups

---

## Responsive Design

### Adjust during collection:

Monitor → Analyze → Adapt

- Change cluster sizes
- Adjust incentives
- Modify protocols

**Bottom line:** Real-time optimization

---

## Case Study: DHS

```{r dhs_case}
# Demographic Health Survey
dhs <- data.frame(
  Country = c("Kenya", "Nigeria", "Ethiopia"),
  PSUs = c(400, 600, 450),
  HH_per_PSU = c(30, 25, 28),
  Total = c(12000, 15000, 12600),
  DEFF = c(2.1, 2.3, 2.2)
)

kable(dhs) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Standard design works globally

---

## Case Study: LSMS

```{r lsms_case}
# Living Standards Measurement
lsms <- data.frame(
  Component = c("PSUs", "Panel", "Waves", "Topics"),
  Detail = c("200-300", "Yes", "3-4", "Comprehensive")
)

kable(lsms) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Complex but valuable

---

## Practical Tips

### From experience:

1. **Start simple** - complicate gradually
2. **Document everything** - future you needs it
3. **Check weights** - always
4. **Monitor quality** - continuously
5. **Have backups** - PSUs will fail

**Bottom line:** Experience teaches

---

## Common Pitfalls

### Avoid these:

❌ Ignoring ICC in planning  
❌ Too few PSUs (< 30)  
❌ Extreme cluster sizes  
❌ No replacement PSUs  
❌ Wrong variance formula  

**Bottom line:** Learn from mistakes

---

## Decision Tree

```{r decision_tree}
cat("Cluster Sampling Decision:

Population clustered? → Yes → Continue
                    ↓
                    No → Use SRS/Stratified

Complete frame? → No → Must cluster
              ↓
              Yes → Consider costs

High travel costs? → Yes → Cluster
                 ↓
                 No → Maybe SRS")
```

**Bottom line:** Logic guides choice

---

## Group Exercise

### Discuss (3 minutes):

Design a survey for your country:
- 10,000 household sample
- Provincial estimates needed
- Budget: $500,000
- Timeline: 3 months

What's your design?

**Bottom line:** Apply knowledge

---

## Innovation Corner

### Emerging approaches:

- **GPS-based sampling**
- **Satellite imagery frames**
- **Mobile phone surveys**
- **Hybrid designs**

**Bottom line:** Field evolving

---

## Climate Considerations

```{r climate_cluster}
# Environmental impact
impact <- data.frame(
  Design = c("SRS", "Clustered"),
  Travel_km = c(50000, 15000),
  CO2_tons = c(5.0, 1.5),
  Reduction = c("0%", "70%")
)

kable(impact) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Clustering greener

---

## Part 3 Summary

### You've learned:

✅ Optimal cluster sizes  
✅ Cost-variance trade-offs  
✅ Advanced variance estimation  
✅ Complex design handling  
✅ Real-world applications  

**Bottom line:** Optimization mastered!

---

## Key Takeaways

```{r key_takeaways}
takeaways <- data.frame(
  Number = 1:5,
  Lesson = c("Optimize cluster size for cost/variance",
            "Use appropriate variance method",
            "Plan for non-response",
            "Document thoroughly",
            "Software handles complexity")
)

kable(takeaways) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Core lessons to remember

---

## Quick Review

### Fill in blanks:

1. Optimal m = √[___________]
2. DEFF = 1 + _______
3. Ultimate cluster uses _______ totals
4. Higher ICC → _______ clusters

**Bottom line:** Test your knowledge

---

## Break Announcement

## ☕ 15-Minute Break

### Coming in Part 4:
- Complex case studies
- Software workshop
- Troubleshooting

### Challenge:
Calculate DEFF for your last survey

**Bottom line:** Rest and return!

---

class: center, middle, inverse

# End of Part 3

## Slides 151-225 Complete

### Next: Part 4 - Advanced Applications

---

---

## Welcome to Part 4!

### Advanced Applications & Practice

Topics for this section:
- Complex case studies
- Software workshop
- Field challenges
- Quality assurance
- Practical exercises

**Bottom line:** Real-world application focus

---

## Case Study: National Health Survey

```{r health_survey_design}
# Real national health survey design
health_design <- data.frame(
  Parameter = c("Target Population", "Sample Size", "PSUs", 
                "Stratification", "Stages", "Domains"),
  Value = c("15M adults", "15,000", "600 EAs",
            "Province × Urban/Rural", "2", "12 provinces")
)

kable(health_design) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Large-scale implementation

---

## Design Decisions

```{r design_decisions}
# Key decisions made
decisions <- data.frame(
  Decision = c("PSU Definition", "Selection Method", 
               "Cluster Size", "Replacement"),
  Choice = c("Census EAs", "PPS Systematic", 
             "25 households", "2 per PSU"),
  Rationale = c("Available frame", "Size variation",
                "Cost-variance balance", "Non-response")
)

kable(decisions) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Each choice has reasons

---

## Sampling Frame Preparation

```{r frame_prep}
# Frame preparation steps
cat("Frame Preparation Process:

1. Obtain census EA list
2. Update population counts
3. Merge/split small/large EAs
4. Assign to strata
5. Calculate size measures
6. Sort for implicit stratification
7. Check for duplicates
8. Document changes")
```

**Bottom line:** Frame quality critical

---

## Size Measure Updates

```{r size_updates, fig.height=3.5}
# Population growth adjustments
years_since_census <- 0:10
growth_rate <- 0.025
adjustment_factor <- (1 + growth_rate)^years_since_census

data.frame(Years = years_since_census, 
           Factor = adjustment_factor) %>%
  ggplot(aes(x = Years, y = Factor)) +
  geom_line(size = 1.5, color = "darkgreen") +
  geom_point(size = 2) +
  theme_minimal() +
  labs(y = "Adjustment Factor",
       title = "Population Growth Adjustments")
```

**Bottom line:** 28% growth over 10 years

---

## Stratification Scheme

```{r stratification_scheme}
# Complex stratification
strata_def <- data.frame(
  Province = rep(c("North", "South", "East"), each = 2),
  Area = rep(c("Urban", "Rural"), 3),
  Stratum_ID = 1:6,
  PSUs_Total = c(80, 120, 100, 150, 70, 80),
  PSUs_Sample = c(40, 60, 50, 75, 35, 40)
)

kable(strata_def) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** 6 explicit strata

---

## Sample Allocation

```{r sample_allocation, fig.height=3.5}
# Allocation across strata
strata_def %>%
  ggplot(aes(x = interaction(Province, Area), 
             y = PSUs_Sample, fill = Area)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Stratum", y = "Number of PSUs",
       title = "PSU Allocation by Stratum")
```

**Bottom line:** Proportional to size allocation

---

## Field Organization

```{r field_org}
# Field team structure
teams <- data.frame(
  Level = c("National", "Regional", "Team", "Interviewer"),
  Count = c(1, 4, 20, 80),
  Responsibility = c("Overall coordination",
                    "Regional supervision",
                    "PSU supervision",
                    "Data collection")
)

kable(teams) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Hierarchical supervision

---

## Timeline Planning

```{r timeline}
# Project timeline
timeline <- data.frame(
  Month = 1:6,
  Activity = c("Design & Training", "Listing", 
               "Data Collection 1", "Data Collection 2",
               "Data Processing", "Report Writing")
)

kable(timeline) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** 6-month implementation

---

## Budget Breakdown

```{r budget_breakdown, fig.height=3.5}
# Cost components
budget <- data.frame(
  Component = c("Personnel", "Travel", "Equipment", 
                "Training", "Other"),
  Amount = c(200000, 150000, 50000, 30000, 70000),
  Percent = c(40, 30, 10, 6, 14)
)

ggplot(budget, aes(x = "", y = Percent, fill = Component)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_minimal() +
  labs(title = "Budget Distribution")
```

**Bottom line:** 70% for personnel and travel

---

## Quality Control Measures

```{r quality_control}
# Quality assurance steps
qa_measures <- data.frame(
  Stage = c("Training", "Listing", "Interview", "Data Entry"),
  Measure = c("Competency test", "10% verification",
              "5% re-interview", "Double entry 10%"),
  Target = c(">80% pass", "<5% error", 
             ">95% match", "<1% error")
)

kable(qa_measures) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Quality at every stage

---

## Non-Response Protocol

```{r nonresponse_protocol}
# Handling non-response
protocol <- data.frame(
  Visit = c(1, 2, 3, "Replacement"),
  Time = c("Initial", "Evening", "Weekend", "New HH"),
  Action = c("First attempt", "Different time",
             "Different day", "Use replacement")
)

kable(protocol) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Three attempts before replacement

---

## Weight Calculation Steps

```{r weight_calc_detailed}
cat("Weight Calculation Process:

1. Base weight = 1/(P1 × P2)
2. Non-response adjustment
3. Post-stratification
4. Trimming extremes
5. Final normalization

Document each step!")
```

**Bottom line:** Multiple adjustments needed

---

## Weight Components Example

```{r weight_components_ex}
# Detailed weight calculation
weight_calc <- data.frame(
  Component = c("PSU selection", "HH selection", 
                "Non-response", "Post-strat", "Final"),
  Probability = c(0.05, 0.04, 0.85, NA, NA),
  Weight_Factor = c(20, 25, 1.18, 0.95, 1.0),
  Cumulative = c(20, 500, 590, 560, 560)
)

kable(weight_calc) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Final weight = 560

---

## Software Workshop: R

```{r r_workshop}
# Complete R workflow
cat("# Load packages
library(survey)
library(sampling)

# Read and prepare frame
frame <- read.csv('psu_frame.csv')

# PPS selection
n_psu <- 100
frame$prob <- n_psu * frame$size / sum(frame$size)
s <- cluster(frame, 'psu_id', n_psu, 'systematic', frame$size)

# Create survey design
design <- svydesign(
  ids = ~psu_id + hh_id,
  strata = ~stratum,
  weights = ~weight,
  data = survey_data
)")
```

**Bottom line:** Complete R implementation

---

## Software Workshop: Stata

```{r stata_workshop}
cat("* Stata implementation
* Load data
use survey_data.dta, clear

* Set survey design
svyset psu_id [pw=weight], ///
  strata(stratum) fpc(fpc1) || ///
  hh_id, fpc(fpc2)

* Estimation
svy: mean income
svy: proportion education_level
svy: total expenditure

* Subpopulation
svy, subpop(if urban==1): mean income")
```

**Bottom line:** Stata handles complex designs

---

## Common Estimation Commands

```{r estimation_commands}
# Key commands comparison
commands <- data.frame(
  Statistic = c("Mean", "Total", "Proportion", "Quantile"),
  R = c("svymean()", "svytotal()", "svymean()", "svyquantile()"),
  Stata = c("svy: mean", "svy: total", "svy: proportion", "svy: _pctile")
)

kable(commands) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Similar across packages

---

## Troubleshooting Guide

```{r troubleshooting}
# Common issues and solutions
issues <- data.frame(
  Problem = c("Lonely PSU", "Missing weights", "Memory error"),
  Solution = c("options(survey.lonely.psu='adjust')",
               "Check weight calculation",
               "Use subset or database")
)

kable(issues) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Know common fixes

---

## Field Challenge: Access

```{r access_challenge}
# Dealing with inaccessible PSUs
access_solutions <- data.frame(
  Issue = c("Security", "Geographic", "Refusal", "Weather"),
  Solution = c("Use replacement", "Alternative transport",
               "Community engagement", "Reschedule"),
  Impact = c("Bias risk", "Cost increase", 
             "Timeline delay", "Timeline delay")
)

kable(access_solutions) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Plan for challenges

---

## Field Challenge: Listing

```{r listing_challenge, fig.height=3.5}
# Listing quality issues
listing_issues <- data.frame(
  Issue = c("Boundary\nunclear", "Rapid\ngrowth", 
            "Seasonal\nmigration", "Incomplete\nmaps"),
  Frequency = c(25, 35, 20, 20)
)

ggplot(listing_issues, aes(x = Issue, y = Frequency)) +
  geom_bar(stat = "identity", fill = "coral") +
  theme_minimal() +
  labs(y = "% of PSUs affected",
       title = "Common Listing Challenges")
```

**Bottom line:** 35% have rapid growth issues

---

## Technology Integration

```{r technology}
# Modern survey tools
tech_tools <- data.frame(
  Tool = c("Tablets", "GPS", "Cloud sync", "Real-time monitoring"),
  Benefit = c("Data quality", "PSU verification", 
              "Immediate backup", "Quality control"),
  Challenge = c("Training", "Accuracy", "Connectivity", "Dashboard setup")
)

kable(tech_tools) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Technology improves quality

---

## Exercise: Complete Design

### Your task (10 minutes):

Design a two-stage cluster sample:
- Population: 5 million
- Target sample: 8,000 households  
- Budget: $400,000
- Need provincial estimates

Calculate:
1. Number of PSUs
2. Cluster size
3. Stratification
4. Weights

**Bottom line:** Apply all concepts

---

## Exercise Solution Part 1

```{r exercise_solution_1}
# Given parameters
population <- 5000000
target_n <- 8000
budget <- 400000
provinces <- 8

# Cost assumptions
c_fixed <- 50000
c_psu <- 1200
c_hh <- 35

# Available for sampling
available <- budget - c_fixed
# For 8000 households
```

**Bottom line:** $350,000 for fieldwork

---

## Exercise Solution Part 2

```{r exercise_solution_2}
# Optimal cluster size (assume ICC = 0.05)
icc <- 0.05
m_opt <- sqrt(c_psu * (1 - icc) / (c_hh * icc))
m_use <- 25  # Practical choice

# Number of PSUs
n_psu <- 8000 / m_use

# Allocation to provinces
psu_per_province <- n_psu / provinces

c(Total_PSUs = n_psu,
  Per_Province = psu_per_province,
  Cluster_Size = m_use)
```

**Bottom line:** 320 PSUs, 40 per province

---

## Exercise Solution Part 3

```{r exercise_solution_3}
# Weight calculation example
# Province 1, PSU 1, HH 1
N_prov <- 625000  # Province population
n_psu_prov <- 40  # PSUs in province
N_psu <- 2500     # PSU population
n_hh <- 25        # HH selected

# Probabilities
p1 <- n_psu_prov / (625000/2500)  # PSU selection
p2 <- n_hh / 100                   # HH selection (assume 100 HH per PSU)

weight <- 1 / (p1 * p2)

c(PSU_Prob = round(p1, 4),
  HH_Prob = p2,
  Weight = round(weight))
```

**Bottom line:** Weight = 250 for this household

---

## Variance Estimation Practice

```{r variance_practice}
# Using survey package
cat("# Create design object
design <- svydesign(
  ids = ~psu_id,
  strata = ~province,
  weights = ~weight,
  data = your_data
)

# Estimate with correct SE
result <- svymean(~income, design)
confint(result)  # 95% CI")
```

**Bottom line:** Let software handle variance

---

## Report Template

```{r report_template}
cat("Survey Report Structure:

1. Executive Summary
2. Methodology
   - Design overview
   - Sampling frame
   - Selection procedures
   - Weight calculation
3. Quality Indicators
   - Response rates
   - Design effects
   - Coefficient of variation
4. Key Findings
5. Appendices
   - Detailed tables
   - Questionnaire")
```

**Bottom line:** Standard reporting format

---

## Quality Indicators Report

```{r quality_report}
# Standard quality metrics
quality_report <- data.frame(
  Indicator = c("Response Rate", "DEFF", "CV", 
                "Item Non-response", "ICC"),
  Overall = c("82%", "2.1", "2.5%", "3%", "0.06"),
  Urban = c("78%", "1.9", "2.8%", "2%", "0.04"),
  Rural = c("85%", "2.3", "3.1%", "4%", "0.08")
)

kable(quality_report) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Document quality by domain

---

## Lessons from the Field

### Experienced surveyor tips:

1. **Always have Plan B** - PSUs will fail
2. **Train for reality** - not just theory
3. **Pilot thoroughly** - find problems early
4. **Communicate constantly** - with teams
5. **Document deviations** - transparency

**Bottom line:** Experience matters

---

## Regional Variations

```{r regional_var, fig.height=3.5}
# Response patterns by region
regions <- data.frame(
  Region = c("North", "South", "East", "West", "Central"),
  Response_Rate = c(85, 78, 82, 75, 88),
  Avg_Attempts = c(1.8, 2.3, 2.0, 2.5, 1.6)
)

ggplot(regions, aes(x = Avg_Attempts, y = Response_Rate)) +
  geom_point(size = 4, color = "darkblue") +
  geom_text(aes(label = Region), vjust = -1) +
  theme_minimal() +
  labs(x = "Average Contact Attempts",
       y = "Response Rate (%)",
       title = "Regional Response Patterns")
```

**Bottom line:** Adjust effort by region

---

## Cost Tracking

```{r cost_tracking}
# Monitor costs during fieldwork
cost_monitor <- data.frame(
  Week = 1:4,
  Planned = c(25, 50, 75, 100),
  Actual = c(22, 48, 77, 95),
  Status = c("On track", "On track", "Slight over", "Under budget")
)

kable(cost_monitor) %>%
  kable_styling(font_size = 12)
```

**Bottom line:** Track weekly

---

## Group Discussion

### Share experiences (5 minutes):

In groups of 3-4, discuss:

1. Biggest cluster sampling challenge?
2. How did you solve it?
3. What would you do differently?

Report back one key insight

**Bottom line:** Learn from peers

---

## Innovation Spotlight

```{r innovation}
# Emerging methods
innovations <- data.frame(
  Method = c("Adaptive sampling", "GIS integration", 
             "Machine learning weights", "Real-time dashboards"),
  Application = c("Rare populations", "Frame updates",
                 "Non-response adjustment", "Quality monitoring"),
  Maturity = c("Research", "Operational", "Pilot", "Operational")
)

kable(innovations) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Field advancing rapidly

---

## Ethical Considerations

```{r ethics}
cat("Ethical Guidelines for Cluster Sampling:

1. Community consent before PSU entry
2. Respect local customs and leaders  
3. Ensure respondent anonymity
4. Avoid cluster identification in reports
5. Fair representation of all groups
6. Transparent selection process
7. Return findings to communities")
```

**Bottom line:** Ethics essential

---

## Environmental Impact

```{r environmental, fig.height=3.5}
# Carbon footprint comparison
carbon <- data.frame(
  Design = c("SRS", "Cluster", "Two-stage", "Stratified\nCluster"),
  CO2_kg = c(5000, 2000, 1800, 1500)
)

ggplot(carbon, aes(x = Design, y = CO2_kg)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
  theme_minimal() +
  labs(y = "CO2 Emissions (kg)",
       title = "Environmental Impact by Design")
```

**Bottom line:** Clustering reduces emissions 70%

---

## Future Directions

### Where field is heading:

- **Hybrid frames** (area + list)
- **Mixed modes** (F2F + phone + web)
- **Continuous surveys** (rolling samples)
- **Small area estimation** (model-based)
- **Paradata utilization** (process data)

**Bottom line:** Complexity increasing

---

## Resources

```{r resources}
resources <- data.frame(
  Type = c("Books", "Software", "Guidelines", "Forums"),
  Resource = c("Lohr (2021), Valliant et al (2018)",
              "R survey, Stata svy, SUDAAN",
              "UN Stats, DHS, LSMS",
              "Survey Stats listserv, AAPOR")
)

kable(resources) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Continue learning

---

## Certification Prep

### Key concepts for exams:

☐ ICC calculation and interpretation  
☐ DEFF formula and application  
☐ PPS selection procedures  
☐ Weight calculation  
☐ Variance estimation methods  
☐ Software implementation  

**Bottom line:** Master fundamentals

---

## Final Exercise

### Comprehensive problem (15 minutes):

A country needs health survey:
- 10M population in 5 regions
- Varies by urbanization (30% urban)
- Need estimates by region and urban/rural
- Budget: $750,000
- Timeline: 6 months

Design complete sample!

**Bottom line:** Integrate everything

---

## Part 4 Summary

### You've accomplished:

✅ Analyzed complex cases  
✅ Practiced software implementation  
✅ Solved field challenges  
✅ Designed complete surveys  
✅ Understood quality measures  

**Bottom line:** Ready for practice!

---

## Day 3 Reflection

### Consider:

1. Most valuable concept learned?
2. How will you apply this?
3. What needs more practice?
4. Questions remaining?

Write brief reflection

**Bottom line:** Consolidate learning

---

## Tomorrow Preview

### Day 4: Complex Designs

- Combining techniques
- Multi-phase sampling  
- Panel surveys
- Rotating samples
- Advanced estimation

**Bottom line:** Building on today

---

## Break Time!

## ☕ 15-Minute Break

### Before Part 5:
- Submit exercises
- Prepare questions
- Stretch and hydrate

### Final session:
Case studies and Q&A

**Bottom line:** Almost done!

---

class: center, middle, inverse

# End of Part 4

## Slides 226-300 Complete

### Next: Part 5 - Case Studies & Wrap-up

---


---

## Welcome to Part 5!

### Final Session: Case Studies & Synthesis

Topics for final 50 slides:
- Comprehensive case studies
- Integration of concepts
- Q&A session
- Action planning
- Day 3 wrap-up

**Bottom line:** Bringing it all together

---

## Case Study 1: Agricultural Survey

```{r agri_survey}
# National agricultural survey design
agri_design <- data.frame(
  Component = c("Population", "Frame", "Stratification",
                "PSUs", "Sample Size", "Stages"),
  Detail = c("2M farms", "Agricultural census", 
             "Region × Farm size", "800 EAs",
             "10,000 farms", "2-stage")
)

kable(agri_design) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Complex agricultural design

---

## Agricultural Design Details

```{r agri_details}
# Specific design features
agri_features <- data.frame(
  Stage = c("Stage 1", "Stage 2"),
  Unit = c("Enumeration Area", "Farm"),
  Method = c("PPS by cultivated area", "Stratified by size"),
  Number = c("800 EAs", "12-15 farms/EA"),
  Challenge = c("Outdated frame", "Seasonal variation")
)

kable(agri_features) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Size-based selection crucial

---

## Agricultural ICC Values

```{r agri_icc, fig.height=3.5}
# Typical ICC for agricultural variables
agri_vars <- data.frame(
  Variable = c("Crop yield", "Farm income", "Livestock", 
               "Technology", "Labor"),
  ICC = c(0.15, 0.12, 0.18, 0.08, 0.05)
)

ggplot(agri_vars, aes(x = reorder(Variable, ICC), y = ICC)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  labs(x = "", y = "Intraclass Correlation",
       title = "Agricultural Variables Show High Clustering")
```

**Bottom line:** High ICCs require careful design

---

## Case Study 2: Education Survey

```{r education_survey}
# School-based survey design
edu_design <- data.frame(
  Level = c("National", "Districts", "Schools", "Students"),
  Count = c("1", "50", "300", "9,000"),
  Selection = c("Census", "All", "PPS by enrollment", "SRS within class"),
  Weight_Range = c("1", "1", "5-50", "100-500")
)

kable(edu_design) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Three-stage hierarchical design

---

## School Clustering Effects

```{r school_clustering}
# Design effects in education
edu_deff <- data.frame(
  Outcome = c("Test scores", "Attendance", "Resources", "SES"),
  ICC = c(0.20, 0.15, 0.35, 0.25),
  Class_Size = c(30, 30, 30, 30),
  DEFF = c(6.8, 5.4, 11.2, 8.3)
)

kable(edu_deff) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** School effects very strong

---

## Case Study 3: Business Survey

```{r business_survey}
# Establishment survey design
business_design <- data.frame(
  Stratum = c("Large", "Medium", "Small", "Micro"),
  Frame_Count = c(500, 2000, 10000, 50000),
  Sample = c(500, 400, 500, 200),
  Method = c("Census", "SRS", "Cluster by area", "Area sample")
)

kable(business_design) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Mixed approach by size

---

## Business Survey Challenges

```{r business_challenges}
cat("Unique Business Survey Issues:

1. Frame deterioration (20% annual change)
2. Multi-establishment enterprises
3. Seasonal businesses
4. Informal sector coverage
5. Size measure volatility
6. Industry classification changes
7. Non-response correlation with size

Solutions: Adaptive design, multiple frames")
```

**Bottom line:** Business surveys complex

---

## Case Study 4: Panel Survey

```{r panel_survey}
# Longitudinal design
panel_design <- data.frame(
  Wave = 1:5,
  Year = 2020:2024,
  Original_PSUs = 200,
  Original_HH = 4000,
  Retained = c(4000, 3600, 3200, 2850, 2500),
  Attrition = c("0%", "10%", "20%", "29%", "38%")
)

kable(panel_design) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Attrition major challenge

---

## Panel Attrition Analysis

```{r panel_attrition, fig.height=3.5}
# Attrition patterns
waves <- 1:5
retained <- c(100, 90, 80, 71, 62)

data.frame(Wave = waves, Retained = retained) %>%
  ggplot(aes(x = Wave, y = Retained)) +
  geom_line(size = 1.5, color = "red") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(y = "% Retained",
       title = "Cumulative Panel Attrition")
```

**Bottom line:** 38% loss over 5 waves

---

## Case Study 5: Rapid Assessment

```{r rapid_assessment}
# Emergency survey design
rapid_design <- data.frame(
  Parameter = c("Timeline", "Sample", "Method", 
                "PSUs", "Technology", "Quality"),
  Standard = c("6 months", "10,000", "Face-to-face",
              "400", "Paper", "Full QA"),
  Rapid = c("3 weeks", "1,200", "Phone + F2F",
           "30", "Mobile", "Minimal QA")
)

kable(rapid_design) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Speed vs quality trade-off

---

## Integration Exercise

### Combine all concepts (10 minutes):

Design a survey with:
- Stratification (urban/rural)
- Two-stage cluster sampling
- PPS selection
- Panel component
- Domain requirements

Work in pairs!

**Bottom line:** Synthesize learning

---

## Integration Solution

```{r integration_solution}
# Integrated design solution
integrated <- data.frame(
  Component = c("Strata", "PSUs/stratum", "HH/PSU", 
                "Panel %", "Domains", "Total n"),
  Urban = c("4 regions", "30", "20", "50%", "4", "2,400"),
  Rural = c("4 regions", "50", "20", "50%", "4", "4,000"),
  Total = c("8 strata", "320", "20", "50%", "8", "6,400")
)

kable(integrated) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Complex but feasible

---

## Weight Calculation Review

```{r weight_review}
# Complete weight calculation
cat("Final Weight Components:

W = W_base × W_nr × W_ps × W_trim

Where:
W_base = 1/(P1 × P2)
W_nr = 1/response_rate
W_ps = population/weighted_sum
W_trim = trimming_factor

Always verify: Σ(weights) ≈ Population")
```

**Bottom line:** Multiple adjustments

---

## Variance Estimation Summary

```{r variance_summary}
# Methods comparison
var_methods <- data.frame(
  Method = c("Taylor", "Jackknife", "Bootstrap", "BRR"),
  Best_For = c("Means/totals", "Complex stats", 
               "Any statistic", "2-PSU strata"),
  Speed = c("Fastest", "Medium", "Slowest", "Fast"),
  Implementation = c("Easy", "Medium", "Complex", "Medium")
)

kable(var_methods) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Choose appropriate method

---

## Software Comparison

```{r software_compare}
# Survey software capabilities
software <- data.frame(
  Feature = c("Complex designs", "Graphics", "Speed", 
              "Documentation", "Cost"),
  R = c("Excellent", "Excellent", "Good", "Good", "Free"),
  Stata = c("Excellent", "Good", "Excellent", "Excellent", "$$$"),
  SAS = c("Excellent", "Fair", "Excellent", "Good", "$$$$"),
  SPSS = c("Good", "Good", "Good", "Fair", "$$$")
)

kable(software) %>%
  kable_styling(font_size = 9)
```

**Bottom line:** R and Stata dominate

---

## Quality Assurance Checklist

```{r qa_checklist}
cat("✓ Frame coverage checked
✓ PSU selection documented
✓ Weights calculated correctly
✓ Variance method appropriate
✓ Response rates monitored
✓ Data quality indicators computed
✓ Documentation complete
✓ Reproducible code archived
✓ Metadata preserved
✓ Reports peer-reviewed")
```

**Bottom line:** Systematic QA essential

---

## Common Mistakes to Avoid

```{r mistakes}
# Top 10 mistakes
mistakes <- data.frame(
  Rank = 1:5,
  Mistake = c("Ignoring clustering in analysis",
              "Wrong variance formula",
              "No weight documentation",
              "Too few PSUs (<30)",
              "Not planning for non-response"),
  Impact = c("Invalid inference", "Wrong SEs",
             "Non-reproducible", "Unstable estimates",
             "Biased results")
)

kable(mistakes) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Learn from others' errors

---

## Best Practices Summary

### Golden rules:

1. **Document everything** - Future you will thank you
2. **Keep it simple** - Complex enough, not more
3. **Plan for problems** - They will happen
4. **Use software** - Don't calculate by hand
5. **Check, check, check** - Verify at each step

**Bottom line:** Quality through discipline

---

## Advanced Topics Preview

### Where to go next:

- Small area estimation
- Adaptive sampling
- Network sampling  
- Bayesian survey inference
- Machine learning integration
- Real-time sampling

**Bottom line:** Field evolving rapidly

---

## Q&A Session

### Open Floor (20 minutes)

Your questions on:
- Concepts unclear?
- Implementation challenges?
- Software issues?
- Country-specific needs?
- Resources needed?

**Bottom line:** No question too basic

---

## Question 1: Small PSUs

```{r q1_small_psu}
cat("Q: What if PSUs are very small?

A: Options:
1. Combine adjacent PSUs
2. Use larger area units
3. Census small PSUs
4. Adjust design (3-stage)

Key: Minimum ~50-75 households/PSU")
```

**Bottom line:** Flexibility needed

---

## Question 2: Budget Constraints

```{r q2_budget}
cat("Q: Budget cut mid-survey?

A: Strategies:
1. Reduce cluster size (not PSUs!)
2. Subsample domains
3. Phone follow-up
4. Reduce questionnaire
5. Phased approach

Never: Reduce PSUs below 30")
```

**Bottom line:** Maintain design integrity

---

## Question 3: High ICC

```{r q3_high_icc, fig.height=3}
# Impact of very high ICC
icc_range <- seq(0, 0.5, 0.05)
deff_25 <- 1 + 24 * icc_range

data.frame(ICC = icc_range, DEFF = deff_25) %>%
  ggplot(aes(x = ICC, y = DEFF)) +
  geom_line(size = 1.5, color = "red") +
  geom_hline(yintercept = 3, linetype = "dashed") +
  theme_minimal() +
  labs(title = "DEFF Explodes with High ICC (m=25)")
```

**Bottom line:** Reduce cluster size

---

## Question 4: Weight Trimming

```{r q4_weights}
cat("Q: When to trim weights?

A: Guidelines:
- If CV(weights) > 2
- If max/min > 10
- If few extreme weights

Methods:
- Trim at percentiles
- Winsorize
- Cap at threshold

Document impact on estimates!")
```

**Bottom line:** Balance bias vs variance

---

## Question 5: Software Choice

```{r q5_software}
# Decision tree for software
cat("Choose Software:

Free? → Yes → R (survey package)
      ↓
      No → Stata usually best

Complex? → Yes → R or Stata
        ↓
        No → Any package

Team skills? → R/Python → R
           ↓
           → Statistical → Stata
           → Business → SPSS/SAS")
```

**Bottom line:** Consider context

---

## Action Planning

### Your next steps:

☐ Review your current design  
☐ Calculate ICC for key variables  
☐ Assess DEFF impact  
☐ Consider design improvements  
☐ Update documentation  
☐ Train team on concepts  

**Bottom line:** Apply immediately

---

## Country-Specific Planning

```{r country_planning}
# Template for country planning
planning <- data.frame(
  Task = c("Assess current practice", "Identify gaps",
           "Design improvements", "Pilot test",
           "Implementation", "Evaluation"),
  Timeline = c("Week 1-2", "Week 3", "Week 4-6",
              "Month 2", "Month 3-6", "Month 7"),
  Responsible = c("Team lead", "All", "Statistician",
                 "Field team", "All", "External")
)

kable(planning) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Systematic improvement

---

## Resources for Continued Learning

```{r resources_detailed}
cat("Essential Resources:

Books:
- Lohr 'Sampling: Design and Analysis' (2021)
- Levy & Lemeshow 'Sampling of Populations' (2013)

Online:
- UCLA IDRE Survey Analysis
- Penn State STAT 506
- CDC NHANES tutorials

Communities:
- Survey Research Methods Section (ASA)
- AAPOR (Public Opinion)")
```

**Bottom line:** Keep learning

---

## Regional Collaboration

```{r regional_collab}
# SADC collaboration opportunities
collaboration <- data.frame(
  Activity = c("Annual workshop", "Peer review",
               "Joint surveys", "Method sharing"),
  Frequency = c("Yearly", "Ongoing", "Biennial", "Quarterly"),
  Benefit = c("Capacity building", "Quality improvement",
             "Cost sharing", "Standardization")
)

kable(collaboration) %>%
  kable_styling(font_size = 11)
```

**Bottom line:** Stronger together

---

## Key Formulas Reference

```{r formula_reference}
# Quick reference card
formulas <- data.frame(
  Concept = c("ICC", "DEFF", "n_eff", "Optimal m", "Weight"),
  Formula = c("ρ = σ²_b/(σ²_b + σ²_w)",
             "1 + (m-1)ρ",
             "n/DEFF",
             "√[c₁(1-ρ)/(c₂ρ)]",
             "1/(P₁ × P₂)")
)

kable(formulas) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Keep handy

---

## Final Group Exercise

### Design challenge (15 minutes):

Your minister wants:
- "Representative" survey
- Results in 1 month
- $50,000 budget
- National + 10 provinces
- n = 10,000

How do you respond?

**Bottom line:** Reality check

---

## Exercise Reality Check

```{r reality_check}
cat("Reality Response:

'Minister, with $50,000 and 1 month:
- Maximum n ≈ 1,000 (not 10,000)
- National OR 3 provinces (not all 10)
- Phone/web only (not face-to-face)
- Existing frame required
- Limited quality assurance

Recommend: Phased approach or 
increased budget/timeline'")
```

**Bottom line:** Manage expectations

---

## Success Stories

```{r success_stories}
# Regional successes
successes <- data.frame(
  Country = c("South Africa", "Kenya", "Ghana"),
  Survey = c("GHS", "KDHS", "GLSS"),
  Innovation = c("Integrated design", "Mobile collection",
                "Community sampling"),
  Impact = c("Policy change", "Health programs", 
            "Poverty reduction")
)

kable(successes) %>%
  kable_styling(font_size = 10)
```

**Bottom line:** Good design matters

---

## Participant Feedback

### Quick evaluation:

Rate 1-5:
1. Content clarity
2. Practical relevance
3. Pace appropriate
4. Examples helpful
5. Ready to apply

Plus: One thing to improve?

**Bottom line:** Help us improve

---

## Day 3 Key Messages

### Remember:

1. **Clustering inevitable** - embrace it
2. **ICC drives design** - measure it
3. **PPS improves efficiency** - use it
4. **Software handles complexity** - trust it
5. **Documentation critical** - do it

**Bottom line:** Foundation established

---

## Certificate Preparation

```{r certificate}
cat("For SADC Certificate:

Demonstrate understanding of:
✓ Cluster sampling concepts
✓ ICC calculation
✓ DEFF interpretation
✓ PPS selection
✓ Two-stage designs
✓ Weight calculation
✓ Variance estimation
✓ Software implementation")
```

**Bottom line:** Competency achieved

---

## Thank You!

### Day 3 Complete!

You've mastered:
- Multi-stage designs
- Cluster sampling
- PPS methodology
- Complex weights
- Variance estimation

See you tomorrow for Day 4!

**Bottom line:** Well done!

---

## Homework

### Before tomorrow:

1. Calculate ICC for one variable in your data
2. Estimate DEFF for your last survey
3. Review PPS selection steps
4. List three improvements for your design
5. Prepare one question for Day 4

**Bottom line:** Reinforce learning

---

## Tomorrow's Preview

### Day 4: Complex Survey Designs

Morning:
- Combining all techniques
- Multi-phase sampling
- Domain estimation

Afternoon:
- Panel designs
- Rotating samples
- Advanced topics

**Bottom line:** Building complexity

---

## Contact Information

```{r contact}
cat("Stay Connected:

Workshop Materials:
github.com/sadc-survey-workshop

Discussion Forum:
sadc-survey.discourse.group

Email Support:
survey.support@sadc.int

WhatsApp Group:
[QR Code Here]")
```

**Bottom line:** Support continues

---

## Final Thought

### Quote of the day:

> "All models are wrong, but some are useful"
> - George Box

Applied to sampling:

> "All samples are biased, but good designs minimize it"

**Bottom line:** Pursue excellence, accept reality

---

## Group Photo!

### 📸 Capture the moment

Stand by your country flag!

Share on social media:
#SADCSurvey #ClusterSampling #Day3Done

**Bottom line:** Celebrate progress

---

## Safe Travels

### Evening reminders:

- Save all materials
- Backup your work
- Review notes tonight
- Rest well
- Hydrate

See you at 8:00 AM tomorrow!

**Bottom line:** Rest and return ready

---

class: center, middle, inverse

# End of Day 3

## Multi-Stage and Cluster Sampling

### Complete! 

#### Total slides: 350
#### Concepts mastered: ✓
#### Ready for Day 4: ✓

---